---
title: "Cell Type-Based Spatial Clustering Analysis"
author: "Mithun & Rajitha"
format: 
  pdf:
    toc: true
    number-sections: true
editor: visual
execute: 
  warning: false
---

#------------------------------------------------------------------------------ \# Required Libraries #------------------------------------------------------------------------------

```{r setup, message=FALSE, warning=FALSE}
# Load libraries required for spatial data processing, clustering, plotting, and data manipulation
library(SpatialExperiment)  # For handling spatial transcriptomics data
library(spdep)              # For spatial dependence and neighborhood analysis
library(adespatial)         # For constrained hierarchical clustering
library(ggplot2)            # For visualization of spatial data
library(dplyr)              # For data manipulation and transformation
library(tidyr)              # For data reshaping and pivoting
library(tibble)             # For enhanced data frame operations
library(cluster)            # For clustering algorithms
library(igraph)             # For graph-based operations in merging
library(reshape2)           # For reshaping data for visualization
```

#------------------------------------------------------------------------------ \# 1. DATA LOADING AND PREPARATION #------------------------------------------------------------------------------

```{r data_loading}
# Load the SpatialExperiment object containing TNBC data
load("03_TNBC_2018_spe.rds")

# Select a specific patient sample for analysis
sample_number <- "Sample_05"
spe_patient <- spe[, spe$sample_id == sample_number]

# Create a data frame with cell IDs and their corresponding cell types
cell_types_df <- tibble(
  cellID = colnames(spe_patient),
  cell_type = spe_patient$mm  # mm column contains cell phenotype information
)

# Display unique cell types for verification and documentation
print("Unique cell types in the dataset:")
unique_types <- unique(cell_types_df$cell_type)
print(unique_types)

# Convert cell types to binary feature matrix using one-hot encoding
feature_mat <- cell_types_df %>%
  mutate(present = 1) %>%  # Mark presence with 1
  pivot_wider(
    id_cols = cellID,
    names_from = cell_type,
    values_from = present,
    values_fill = list(present = 0)  # Mark absence with 0
  )

# Extract spatial coordinates for each cell
coords_mat <- as.data.frame(spatialCoords(spe_patient)) %>%
  rownames_to_column(var = "cellID")
colnames(coords_mat)[2:3] <- c("X", "Y")  # Standardize coordinate column names

# Define feature columns (exclude cellID)
feature_cols <- setdiff(names(feature_mat), "cellID")

#Create dataframe
patient_data <- coords_mat %>%
  select(cellID, X, Y) %>%
  left_join(feature_mat, by = "cellID") %>%
  select(X, Y, all_of(feature_cols), cellID)  #  X, Y first, then features, then cellID

# Prepare matrices for distance calculation and clustering
feature_mat_for_dist <- feature_mat %>% column_to_rownames(var = "cellID")
coords_mat_for_dist <- coords_mat %>% column_to_rownames(var = "cellID")

# Calculate binary distance matrix between cells based on cell type
dist_features <- dist(feature_mat_for_dist, method = "binary")
```

#------------------------------------------------------------------------------ \# 2. PARAMETER SELECTION AND CLUSTERING #------------------------------------------------------------------------------

```{r}
# Define parameter ranges to test for optimal clustering
unique_types <- unique(cell_types_df$cell_type)
num_cell_types <- length(unique_types)
  
  # Set cluster range based on cell types (with min of 4)
min_clusters <- 4
max_clusters <- min(16, num_cell_types)  # Cap at 16 or number of cell types, whichever is smaller
num_clusters <- min_clusters:max_clusters
  
num_neighbors <- c(4:8)  # Testing 4 to 8 neighbors

# Evaluate clustering quality using modified silhouette scores
source("evaluate_silhouette.R")
avg_sils <- evaluate_modified_silhouette(
  num_clusters, num_neighbors, "ward.D2", 
  dist_features, coords_mat_for_dist, plot = FALSE
)
```

```{r parameter_tuning}
# Find parameters that yield the highest average modified silhouette score
optimal_params <- avg_sils %>% 
  arrange(desc(avg_modified_sil)) %>% 
  slice(1)
print(paste("Optimal parameters: k =", optimal_params$num_clusters, 
            "neighbors =", optimal_params$num_neighbors))

# Use the identified optimal parameters for clustering
knn <- optimal_params$num_neighbors
k <- optimal_params$num_clusters

# Build spatial neighborhood structure for constrained clustering
nb <- knn2nb(knearneigh(coords_mat_for_dist, k = knn)) %>% nb2listw(style = "W")
neighbors <- listw2sn(nb)[, 1:2]  # Convert to simple neighbor list format

# Perform constrained hierarchical clustering
constr.clust <- constr.hclust(
  d = dist_features,
  method = "ward.D2",  # Ward's method minimizes within-cluster variance
  links = neighbors,   # Spatial neighborhood constraints
  coords = coords_mat_for_dist
)

# Cut dendrogram to obtain the optimal number of clusters
patient_data$region <- as.factor(cutree(constr.clust, k = k))

# Report cluster sizes for evaluation
cluster_sizes <- table(patient_data$region)
print("Cluster sizes:")
print(cluster_sizes)
```

#------------------------------------------------------------------------------ \# 3. CLUSTER ANALYSIS AND VISUALIZATION #------------------------------------------------------------------------------

```{r cluster_visualization}
# Calculate cell type composition of each identified cluster
cluster_composition <- patient_data %>%
  select(region, all_of(feature_cols)) %>%
  group_by(region) %>%
  summarize(across(everything(), sum)) %>%  # Count cells of each type
  mutate(total = rowSums(across(where(is.numeric)))) %>%  # Calculate total cells
  mutate(across(where(is.numeric) & !matches("total"), ~./total))  # Convert to proportions

# Identify the dominant cell type in each cluster
dominant_cell_type <- cluster_composition %>%
  select(-total) %>%
  pivot_longer(cols = -region, names_to = "cell_type", values_to = "proportion") %>%
  group_by(region) %>%
  slice_max(order_by = proportion, n = 1) %>%  # Get cell type with highest proportion
  select(region, dominant_type = cell_type, proportion)

print(dominant_cell_type)

# Define a colorblind-friendly color palette with enough colors for all clusters
cb_palette_extended <- c("#E69F00", "#CC79A7", "#56B4E9", "#009E73", 
                         "#0072B2", "#D55E00", "#F0E442", "#999999", 
                         "#882255", "#661100","#661", "#009",
                         "#087", "#F40","#C20","#30F")

# Visualize the spatial distribution of clusters
g_regions <- ggplot(patient_data, aes(x = X, y = Y, color = region)) +
  geom_point() +
  scale_color_manual(values = cb_palette_extended) +
  coord_equal() +  # Maintain aspect ratio for accurate spatial representation
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12)
  )

print(g_regions)
```

#------------------------------------------------------------------------------ \# 4. BLOCK CREATION - SPATIAL BASED (PROPOSED METHOD) #------------------------------------------------------------------------------

```{r block_creation}
# Source the block creation function
source("create_blocks.R")

# Number of spatial dimensions (X, Y) to use for block creation
num_spatial_dims <- 2

# Create blocks with appropriate size constraints
blocked_data <- create_blocks(patient_data, num_spatial_dims, knn)

# Print the result structure for verification
print(paste("Created", length(unique(blocked_data$polygon_id)), "blocks across", 
            length(unique(blocked_data$region)), "regions"))
```

```{r}
#------------------------------------------------------------------------------ 
# VALIDATION OF THEORETICAL REFINEMENTS
#------------------------------------------------------------------------------

# 1. Validate heterogeneity-based modulation
heterogeneity_validation <- blocked_data %>%
  group_by(region) %>%
  summarize(
    num_blocks = n_distinct(polygon_id),
    num_cells = n(),
    blocks_per_cell = num_blocks / num_cells * 100  # Normalized measure
  )

# Plot relationship between heterogeneity and block density
ggplot(heterogeneity_validation, 
       aes(x = as.numeric(as.factor(region)), y = blocks_per_cell)) +
  geom_bar(stat = "identity") +
  labs(title = "Block Density by Region",
       subtitle = "Higher values indicate more blocks relative to region size",
       x = "Region", y = "Blocks per 100 cells") +
  theme_minimal()

# 2. Validate power optimization by block size distribution
block_size_dist <- blocked_data %>%
  group_by(polygon_id) %>%
  summarize(size = n()) %>%
  ungroup()

# Plot histogram of block sizes
ggplot(block_size_dist, aes(x = size)) +
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black") +
  geom_vline(xintercept = 20, color = "red", linetype = "dashed") +
  annotate("text", x = 20, y = max(table(cut(block_size_dist$size, breaks = 20)))/2, 
           label = "Minimum for\nstatistical power", hjust = -0.1) +
  labs(title = "Distribution of Block Sizes",
       subtitle = "Power-optimized blocks should maintain adequate sample sizes",
       x = "Number of Cells per Block", y = "Frequency") +
  theme_minimal()

```

#------------------------------------------------------------------------------ \# 5. STATISTICAL VALIDATION OF BLOCK CREATION #------------------------------------------------------------------------------

```{r block_validation}
# Check block size distribution for adequate sample sizes
block_size_dist <- blocked_data %>%
  group_by(polygon_id) %>%
  summarize(size = n()) %>%
  ungroup()

print("Block size summary statistics:")
print(summary(block_size_dist$size))

# Check block homogeneity for cell types
block_homogeneity <- blocked_data %>% 
  group_by(polygon_id) %>% 
  summarise(across(all_of(feature_cols), function(x) mean(x != 0))) %>%
  pivot_longer(-polygon_id) %>%
  group_by(polygon_id) %>%
  summarize(
    mean_cell_type_purity = mean(value),
    max_cell_type_purity = max(value)
  )

print("Block homogeneity summary:")
print(summary(block_homogeneity))



# Calculate meaningful feature values for each block
block_centroids <- blocked_data %>%
  group_by(polygon_id) %>%
  summarize(
    X = mean(X),
    Y = mean(Y),
    # Calculate proportion of dominant cell type in each block
    feature_val = {
      type_counts <- colSums(across(all_of(feature_cols)))
      max(type_counts)/sum(type_counts)
    },
    # Store the dominant cell type name
    dominant_type = names(which.max(colSums(across(all_of(feature_cols)))))
  )

# # Visualize blocks by dominant cell type, use when checking blocks
# ggplot() +
#   # Plot cells with block colors
#   geom_point(data = blocked_data, aes(x = X, y = Y, color = polygon_id), alpha = 0.5) +
#   # Add block centroids labeled by dominant cell type
#   geom_label(data = block_centroids, aes(x = X, y = Y, label = dominant_type),
#              size = 3, fontface = "bold") +
#   coord_equal() +
#   labs(title = "Spatial Distribution of Blocks by Dominant Cell Type") +
#   theme_minimal()
```

#------------------------------------------------------------------------------ \# 6. MMD TESTING FOR STATISTICAL CLUSTER SIMILARITY #------------------------------------------------------------------------------

```{r mmd_testing}
# Perform pairwise MMD tests between clusters
source("mmd_tests.R")
test_results <- multiple_comparison(
  blocked_data,
  as.matrix(dist_features),
  kernel = "IMQ",
  kernel_param = 1,
  nperm = 200,
  adj_p = "BH"
)

print("MMD test results:")
print(test_results)


###alternate to pairwise evaluation?
# Identify clusters that are NOT significantly different
similar_clusters <- test_results %>%
  filter(adj_p >= 0.05) %>%
  select(region_1, region_2)

print("Similar clusters (candidates for merging):")
print(similar_clusters)

# Create adjacency matrix from test results
all_regions <- sort(unique(c(test_results$region_1, test_results$region_2)))
adj_matrix <- matrix(0, nrow=length(all_regions), ncol=length(all_regions))
rownames(adj_matrix) <- all_regions
colnames(adj_matrix) <- all_regions

# Fill adjacency matrix - 1 indicates similar regions
for(i in 1:nrow(similar_clusters)) {
  r1 <- as.character(similar_clusters$region_1[i])
  r2 <- as.character(similar_clusters$region_2[i])
  adj_matrix[r1, r2] <- 1
  adj_matrix[r2, r1] <- 1  # Make symmetric
}

# Print cluster similarity relationships
cat("\nCluster similarity relationships:\n")
for(r1 in all_regions) {
  connections <- names(which(adj_matrix[r1,] == 1))
  if(length(connections) > 0) {
    cat(paste0("Cluster ", r1, " is similar to: ", paste(connections, collapse=", "), "\n"))
  } else {
    cat(paste0("Cluster ", r1, " is distinct from all others\n"))
  }
}
```

-   Store the results as a matrix and plot the graph

```{r}
# Source the script that builds the pairwise results graph.
source("evaluate_pairwise_results.R")
#results_matrix <- pairwise_results_to_matrix_og(test_results, plot = TRUE)

# Convert the pairwise test results into an adjacency matrix and plot the corresponding graph.
results_matrix <- pairwise_results_to_matrix(test_results, plot = TRUE)

```

-   Re-partitioning based on the graph

The pairwise graph indicates that certain clusters have similar feature distributions. Merging these clusters can improve interpretability

**Manual Merging:**

Here the decision is made to merge clusters into two groups, "A" and "B", based on prior statistical evidence

```{r}

# Re-calculate clusters from the constrained hierarchical clustering result.
clusters <- cutree(constr.clust, k = 16)


# Clusters that merge togather
clusters[clusters %in% c(1,10,13,16)] <- "A"
clusters[clusters %in% c(2)] <- "B"
clusters[clusters %in% c(3,6,7,8,11,15)] <- "C"
clusters[clusters %in% c(4)] <- "D"
clusters[clusters %in% c(5)] <- "E"
clusters[clusters %in% c(6,3,8,11)] <- "F"
clusters[clusters %in% c(8,3,6,7,11)] <- "G"
clusters[clusters %in% c(9,12,14)] <- "H"
clusters[clusters %in% c(15,3)] <- "I"
clusters[clusters %in% c(16,1)] <- "J"



# Convert the merged cluster labels back into numeric factors.
# This step renumbers the groups to facilitate further analysis.
patient_data$repartitioned_clusters <- as.factor(as.integer(as.factor(clusters)))
```

-   Plotting the re-partitioned clusters

```{r}

# Define a color palette for the new clusters.
cb_palette <- c("#E69F00", "#56B4E9", "#CC79A7", "#009E73", "#0072B2", "#D55E00", "#F0E442")

# Create a plot of the spatial coordinates with colors representing the re-partitioned clusters.
g_repart <- ggplot(patient_data, aes(x = X, y = Y, color = repartitioned_clusters)) +
  geom_point() +
  scale_color_manual(values = cb_palette) +
  coord_equal() +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12)
  )
  # labs(title = "Repartitioned Clusters")  # Optional title.

# Display the re-partitioned clusters plot.
g_repart
```

-   Proportion of tumor cells captured by the cluster that predominantly contains tumor cells.

```{r}

# Count the number of cells of each type within each re-partitioned cluster.
cluster_cell_counts <- patient_data %>%
  count(cellID, repartitioned_clusters) %>% rename(count = n)

# Count the total number of cells of each type across the entire dataset.
cell_counts <- patient_data %>%
  count(cellID) %>% rename(total_cells = n)

# Merge the two count tables to calculate the proportion of each cell type within each cluster.
summary <- left_join(cluster_cell_counts, cell_counts, by = "cellID")
summary['Proportion'] <- summary['count'] / summary['total_cells']
summary
```

-   Saving the ggplots

```{r}
# ----- Saving the Visual Outputs -----

# Save the re-partitioned clusters plot to a PNG file.
ggsave("re_part_05.png", plot = g_repart, width = 7, height = 5, units = "in")


```

-   Saving the pairwise results plot

```{r}
# Save the pairwise results graph as a PNG file.
# Open a PNG device for output with specified dimensions and resolution.
png("pwr_05.png", width = 5, height = 3, units = "in", res = 300)

# Re-run the function to generate the pairwise results graph (this will plot to the PNG device).
results_matrix <- pairwise_results_to_matrix(test_results, plot = TRUE)
results_matrix

# Close the PNG device to finish saving the file.
dev.off()
```

#------------------------------------------------------------------------------ \# 7. 3MT PRESENTATION FIGURES #------------------------------------------------------------------------------

```{r}
#| echo: true
library(gridExtra)
library(ggplot2) # Ensure ggplot2 is loaded for plot modifications

# --- Generate Figures for Slide 2 --- 

# Plot 1: Initial Clusters (Save as initial_clusters_plot.png)
initial_clusters_plot <- ggplot(patient_data, aes(x = X, y = Y, color = region)) +
  geom_point(size=1.0) +
  scale_color_manual(values = cb_palette_extended) + # Use palette defined earlier
  coord_equal() +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(size=14, face="bold", hjust = 0.5),
    axis.text = element_text(size=10),
    axis.title = element_text(size=12)
  ) +
  labs(title = "Initial Clusters")

# Save Plot 1
ggsave("initial_clusters_plot.png", initial_clusters_plot, width = 4.5, height = 4, units = "in", dpi=300)

# Plot 2: Pairwise Similarity Graph (Save as similarity_graph.png)
# Source the script containing the plotting function
source("evaluate_pairwise_results.R") 

# Open a PNG device for the similarity graph
png("similarity_graph.png", width = 4.5, height = 4, units = "in", res = 300)

# Generate the plot directly to the PNG device
pairwise_results_to_matrix(test_results, plot = TRUE)

# Close the PNG device
dev.off()

# --- Generate Figure for Slide 3 --- (Combined plot)

# Plot 3: Repartitioned Clusters (after MMD-based merging)
repartitioned_clusters_plot <- ggplot(patient_data, aes(x = X, y = Y, color = repartitioned_clusters)) +
  geom_point(size=1.0) +
  scale_color_manual(values = cb_palette) + # Use palette defined earlier
  coord_equal() +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(size=14, face="bold", hjust = 0.5),
    axis.text = element_text(size=10),
    axis.title = element_text(size=12)
  ) +
  labs(title = "Repartitioned Clusters") 

# Plot 4: Ground Truth Cell Types
# Create a dataframe with coordinates and cell types directly from the source
cell_types_spatial <- data.frame(
  X = spatialCoords(spe_patient)[,1],
  Y = spatialCoords(spe_patient)[,2],
  cell_type = spe_patient$mm
)

# Cell types plot (no legend for combined figure)
cell_types_plot_no_legend <- ggplot(cell_types_spatial, aes(x = X, y = Y, color = cell_type)) +
  geom_point(size=1.0) +
  # Custom color palette for all cell types
  scale_color_manual(values = c( 
    "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999",
    "#882255", "#661100", "#44AA99", "#332288", "#117733", "#AA4499", "#88CCEE", "#6699CC" 
  )) +
  coord_equal() +
  # Ensure all data points are shown
  scale_x_continuous(limits = c(min(cell_types_spatial$X), max(cell_types_spatial$X))) +
  scale_y_continuous(limits = c(min(cell_types_spatial$Y), max(cell_types_spatial$Y))) +
  theme_minimal() +
  theme(
    legend.position = "none", # No legend in combined plot
    plot.title = element_text(size=14, face="bold", hjust = 0.5),
    axis.text = element_text(size=10),
    axis.title = element_text(size=12)
  ) +
  labs(title = "Cell Types")

# Arrange Slide 3 plots side-by-side
slide3_combined_plot <- grid.arrange(
  repartitioned_clusters_plot, 
  cell_types_plot_no_legend, 
  ncol=2)

# Save Slide 3 figure (matches the filename in the LaTeX file)
ggsave("figure_slide3.png", slide3_combined_plot, width = 9, height = 4, units = "in", dpi=300)
```