------------------------------------------------------------------------

---
title: "TNBC Application with Proust Integration"
author: "Mithun & Rajitha"
format: pdf
editor: visual
execute: 
  warning: false
---

# Required Libraries

```{r}
#| warning: false
library(SpatialExperiment)
library(spdep)
library(adespatial)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
library(cowplot) # For arranging multiple plots
library(parallel)
library(foreach)
library(doParallel)
library(parallelDist)
```

# Data Loading and Preparation

```{r}
# Loading SPE object
load("03_TNBC_2018_spe.rds")

# Re-name the coordinates column
coords <- spatialCoords(spe)
colnames(coords) <- c("X","Y")
spatialCoords(spe) <- coords

# Compartmentalizing cell types
immune_names <- c("B", "CD3 T", "CD4 T", "CD8 T", "DC", "DC/Mono", 
                  "T reg", "NK", "Mono/Neu", "Neu", "Mac", "Other immune")
nonimmune_names <- c("Endothelial", "Epithelial", "Mesenchymal")
tumor_names <- c("Other")

spe$mm[spe$mm %in% immune_names] <- "immune cells"
spe$mm[spe$mm %in%  nonimmune_names] <- "nonimmune cells"
spe$mm[spe$mm %in% tumor_names] <- "tumor cells"
```

# Sample Selection and Feature Extraction

```{r}
# Sample
sample_number = "Sample_26"
patient_id = as.numeric(gsub("Sample_", "", sample_number))  # Extract numeric part
print(paste("Processing sample:", sample_number, "- Patient ID:", patient_id))

# Sub-setting patient data 
spe_patient <- spe[, spe$sample_id == sample_number]

# Creating the feature and coordinates matrix for the selected patient
feature_mat <- as.data.frame(t(assay(spe_patient)))
coords_mat <- as.data.frame(spatialCoords(spe_patient))

# Create a data frame using features and coordinates
patient_data <- cbind(feature_mat, spatialCoords(spe_patient), cells = spe_patient$mm)
```

# Distance Calculation

```{r}
# Removing gene names from the row header
rownames(feature_mat) <- seq_len(nrow(feature_mat))

# Creating distance matrix
dist_features <- dist(feature_mat)
```

# Constrained Hierarchical Clustering section

```{r}
# Parameter Selection using Custom Silhouette Score
source("evaluate_silhouette.R")

# Different neighborhood sizes and number of clusters to calculate modified silhouette
num_clusters <- 4:10
num_neighbors <- c(4, 6, 8)

# Plot and store the modified silhouette score for all combinations
avg_sils <- evaluate_modified_silhouette(num_clusters, num_neighbors, "ward.D2", 
                                        dist_features, coords_mat, plot = FALSE)

# Find parameters that yield the highest average modified silhouette score
optimal_params <- avg_sils %>%
  arrange(desc(avg_modified_sil)) %>%
  slice(1)
print(paste("Optimal parameters: k =", optimal_params$num_clusters, 
            "neighbors =", optimal_params$num_neighbors))

knn <- optimal_params$num_neighbors
k <- optimal_params$num_clusters
```

# Load Proust Clustering Results

```{r}
# Define paths for Proust output files
proust_dir <- "spatial_clusters_for_mmd"
proust_file <- file.path(proust_dir, paste0("patient", patient_id, "_clusters.csv"))

# Try both formats for the hyperparameter file
hyperparams_file <- file.path(proust_dir, paste0("Sample_", patient_id, "_hyperparameters.log"))
hyperparams_file_zero <- file.path(proust_dir, paste0("Sample_", sprintf("%02d", as.numeric(patient_id)), "_hyperparameters.log"))

# Check if Proust output exists
if (file.exists(proust_file)) {
  # Load Proust clustering results
  proust_clusters <- read.csv(proust_file)
  print(paste("Loaded Proust clusters for patient", patient_id))
  print(paste("Found", length(unique(proust_clusters$cluster)), "clusters"))
  
  # Display first few rows
  head(proust_clusters)
  
  # Check both possible hyperparameter file formats
  if (file.exists(hyperparams_file)) {
    hyper_file_to_use <- hyperparams_file
  } else if (file.exists(hyperparams_file_zero)) {
    hyper_file_to_use <- hyperparams_file_zero
  } else {
    hyper_file_to_use <- NULL
    warning(paste("Hyperparameter log file not found at either:", 
                  hyperparams_file, "or", hyperparams_file_zero))
  }
  
  # Load and display hyperparameters if available
  if (!is.null(hyper_file_to_use)) {
    # Read hyperparameter log
    hyper_log <- readLines(hyper_file_to_use)
    
    # Display hyperparameter information
    cat("\nProust Hyperparameters:\n")
    cat("=======================\n")
    cat(paste(hyper_log, collapse = "\n"))
    
    # Extract key parameters for analysis (k_neighbours)
    k_neighbors_line <- grep("k_neighbors:", hyper_log, value = TRUE)
    if (length(k_neighbors_line) > 0) {
      k_neighbors <- as.numeric(gsub(".*k_neighbors: ([0-9]+).*", "\\1", k_neighbors_line))
      print(paste("Using k_neighbors =", k_neighbors))
    }
  }
} else {
  warning(paste("Proust clustering file not found:", proust_file))
  # Create empty dataframe as fallback
  proust_clusters <- data.frame(
    cellLabelInImage = integer(),
    X = numeric(),
    Y = numeric(),
    cluster = integer()
  )
}
```

# Integrate Proust Clusters with Patient Data

```{r}
# Check if we have Proust data
if (nrow(proust_clusters) > 0) {
  # Create a mapping from cell coordinates to Proust cluster
  # We'll need to match cells based on spatial coordinates
  
  # First, create a function to find the nearest cell from Proust results
  find_nearest_cell <- function(x, y, proust_data, max_dist = 5) {
    distances <- sqrt((proust_data$X - x)^2 + (proust_data$Y - y)^2)
    nearest_idx <- which.min(distances)
    if (distances[nearest_idx] <= max_dist) {
      return(proust_data$cluster[nearest_idx])
    } else {
      return(NA)  # No matching cell found within maximum distance
    }
  }
  
  # For larger datasets, we can use a more efficient approach:
  # Create patient_data with coordinates and cell IDs
  patient_coords <- patient_data[, c("X", "Y")]
  
  # For each cell in patient_data, find the nearest Proust cell
  if (nrow(patient_coords) < 10000) { # For smaller datasets
    # Apply the function to each row in patient_data
    patient_data$proust_cluster <- NA
    for (i in 1:nrow(patient_data)) {
      patient_data$proust_cluster[i] <- find_nearest_cell(
        patient_data$X[i], 
        patient_data$Y[i],
        proust_clusters
      )
    }
  } else {
    # For larger datasets, use a more efficient approach with the nearest neighbor algorithm
    library(RANN)
    nn <- nn2(proust_clusters[, c("X", "Y")], patient_coords, k = 1)
    patient_data$proust_cluster <- proust_clusters$cluster[nn$nn.idx[,1]]
    # Set a distance threshold
    patient_data$proust_cluster[nn$nn.dists[,1] > 5] <- NA
  }
  
  # Convert to factor for better plotting
  patient_data$proust_cluster <- as.factor(patient_data$proust_cluster)
  
  # Report on matching results
  n_matched <- sum(!is.na(patient_data$proust_cluster))
  cat(paste0("Successfully matched ", n_matched, " of ", nrow(patient_data), 
            " cells (", round(100 * n_matched / nrow(patient_data), 1), "%)\n"))
} else {
  patient_data$proust_cluster <- NA
  cat("No Proust clusters available for integration\n")
}
```

# Constrained Hierarchical Clustering (Regular Approach)

```{r}
# Creating k-nearest neighborhood
nb <- knn2nb(knearneigh(coords_mat, k = knn)) %>%
  nb2listw(style = "W")

# Creating the data frame needed for the constr.hclust function
neighbors <- listw2sn(nb)[,1:2]

# Constraint hierarchical clustering
constr.clust <- constr.hclust(
  d = dist_features, method = "ward.D2",
  links = neighbors, coords = coords_mat
)

# Combining the optimal number of clusters to the patient_data dataframe
patient_data$region <- as.factor(cutree(constr.clust, k = k))
```

# Visualization of Clustering Results

```{r}
# Define a colorblind-friendly palette
cb_palette <- c("#E69F00", "#CC79A7", "#56B4E9", "#009E73", "#0072B2", 
                "#D55E00", "#F0E442", "#999999", "#FF6666", "#66FF66")

# Plotting standard clustering regions
g_regions <- ggplot(patient_data, aes(x = X, y = Y, color = region)) +
  geom_point(size = 1.5) +
  scale_color_manual(values = cb_palette) +
  coord_equal() +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12)
  ) +
  labs(color = 'Constrained\nClusters', title = "Standard Constrained Clustering")

# Plot cell types
g_cells <- ggplot(patient_data, aes(x = X, y = Y, color = cells)) +
  geom_point(size = 1.5) +
  scale_color_manual(values = cb_palette[1:3]) +
  coord_equal() +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12)
  ) +
  labs(color = 'Cell Type', title = "Cell Type Distribution")

# Plot Proust clusters if available
if (sum(!is.na(patient_data$proust_cluster)) > 0) {
  g_proust <- ggplot(subset(patient_data, !is.na(proust_cluster)), 
                    aes(x = X, y = Y, color = proust_cluster)) +
    geom_point(size = 1.5) +
    scale_color_viridis_d() +
    coord_equal() +
    theme_minimal() +
    theme(
      legend.position = "right",
      axis.text = element_text(size = 12),
      legend.text = element_text(size = 12),
      axis.title = element_text(size = 12),
      legend.title = element_text(size = 12)
    ) +
    labs(color = 'Proust\nCluster', title = "Proust Clustering")
  
  # Arrange all three plots
  plot_grid(g_cells, g_regions, g_proust, ncol = 3)
} else {
  # Just show the two plots if no Proust data
  plot_grid(g_cells, g_regions, ncol = 2)
}
```

# Create Blocks for MMD Testing

```{r}
# Source the script containing the function to create blocks
source("create_blocks.R")

# Create blocks for standard clustering approach
standard_blocks <- create_blocks(patient_data, ncol(feature_mat), knn)

# Create blocks for Proust clustering if available
if (sum(!is.na(patient_data$proust_cluster)) > 0) {
  # Create a copy of patient_data with proust_cluster as region
  proust_patient_data <- patient_data
  proust_patient_data$region <- proust_patient_data$proust_cluster
  
  # Remove rows with NA proust_cluster
  proust_patient_data <- proust_patient_data[!is.na(proust_patient_data$region),]
  
  # Create blocks for Proust clusters
  proust_blocks <- create_blocks(proust_patient_data, ncol(feature_mat), knn)
  
  # Print block statistics
  cat("Standard clustering blocks:", length(unique(standard_blocks$polygon_id)), "\n")
  cat("Proust clustering blocks:", length(unique(proust_blocks$polygon_id)), "\n")
} else {
  cat("Using only standard clustering blocks\n")
}
```

# MMD Testing for Standard Clustering

```{r}
source("mmd_tests.R")
standard_test_results <- multiple_comparison(standard_blocks, as.matrix(dist_features), "IMQ", 1, 200)

# Visualize results
source("evaluate_pairwise_results.R")
standard_results_matrix <- pairwise_results_to_matrix(standard_test_results, plot = TRUE)
```

# MMD Testing for Proust Clustering

```{r}
if (exists("proust_blocks") && nrow(proust_blocks) > 0) {
  # Run MMD tests on Proust clusters
  proust_test_results <- multiple_comparison(proust_blocks, as.matrix(dist_features), "IMQ", 1, 200)
  
  # Visualize results
  proust_results_matrix <- pairwise_results_to_matrix(proust_test_results, plot = TRUE)
  # Compare the number of significant differences
  standard_sig_count <- sum(standard_test_results$p_value < 0.05)
  proust_sig_count <- sum(proust_test_results$p_value < 0.05)
  
  cat("Standard clustering: Found", standard_sig_count, "significant differences out of", 
      nrow(standard_test_results), "comparisons\n")
  cat("Proust clustering: Found", proust_sig_count, "significant differences out of", 
      nrow(proust_test_results), "comparisons\n")
}
```

# Repartitioning Standard Clusters

```{r}
# Original clusters from constrained hierarchical clustering
clusters <- cutree(constr.clust, k = 4)

# Find pairs of clusters with no significant difference
similar_clusters <- standard_test_results %>%
  filter(p_value >= 0.05) %>%
  select(region_1, region_2)

# Merge similar clusters if any exist
if (nrow(similar_clusters) > 0) {
  cat("Merging similar clusters based on MMD results\n")
  
  # Convert to character for merging
  clusters_char <- as.character(clusters)
  
  # For each pair of similar clusters, merge them
  for (i in 1:nrow(similar_clusters)) {
    r1 <- similar_clusters$region_1[i]
    r2 <- similar_clusters$region_2[i]
    cat("Merging clusters", r1, "and", r2, "\n")
    
    # Assign a new label (use the first cluster's name)
    new_label <- paste0("Group_", r1)
    clusters_char[clusters_char %in% c(r1, r2)] <- new_label
  }
  
  # Convert back to factors
  patient_data$merged_clusters <- as.factor(clusters_char)
} else {
  cat("No clusters to merge - all are significantly different\n")
  patient_data$merged_clusters <- as.factor(clusters)
}

# Plotting the re-partitioned clusters
g_repart <- ggplot(patient_data, aes(x = X, y = Y, color = merged_clusters)) +
  geom_point(size = 1.5) +
  scale_color_manual(values = cb_palette) +
  coord_equal() +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 12),
    legend.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12)
  ) +
  labs(color = 'Merged\nClusters', title = "MMD-Guided Re-partitioned Clusters")

print(g_repart)
```

# Repartitioning Proust Clusters (if available)

```{r}
if (exists("proust_test_results") && nrow(proust_test_results) > 0) {
  # Find pairs of Proust clusters with no significant difference
  similar_proust_clusters <- proust_test_results %>%
    filter(p_value >= 0.05) %>%
    select(region_1, region_2)
  
  # Merge similar Proust clusters if any exist
  if (nrow(similar_proust_clusters) > 0) {
    cat("Merging similar Proust clusters based on MMD results\n")
    
    # Make a copy of the original Proust cluster assignment
    proust_merged <- as.character(patient_data$proust_cluster)
    
    # For each pair of similar clusters, merge them
    for (i in 1:nrow(similar_proust_clusters)) {
      r1 <- as.character(similar_proust_clusters$region_1[i])
      r2 <- as.character(similar_proust_clusters$region_2[i])
      cat("Merging Proust clusters", r1, "and", r2, "\n")
      
      # Assign a new label
      new_label <- paste0("P_Group_", r1)
      proust_merged[proust_merged %in% c(r1, r2)] <- new_label
    }
    
    # Convert back to factors
    patient_data$proust_merged <- as.factor(proust_merged)
    
    # Plotting the re-partitioned Proust clusters
    g_proust_repart <- ggplot(subset(patient_data, !is.na(proust_cluster)), 
                             aes(x = X, y = Y, color = proust_merged)) +
      geom_point(size = 1.5) +
      scale_color_viridis_d() +
      coord_equal() +
      theme_minimal() +
      theme(
        legend.position = "right",
        axis.text = element_text(size = 12),
        legend.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        legend.title = element_text(size = 12)
      ) +
      labs(color = 'Merged\nProust Clusters', title = "MMD-Guided Re-partitioned Proust Clusters")
    
    print(g_proust_repart)
  } else {
    cat("No Proust clusters to merge - all are significantly different\n")
    patient_data$proust_merged <- patient_data$proust_cluster
  }
}
```

# Compare Clustering Methods

```{r}
# Calculate the adjusted Rand index (ARI) between clustering methods
if (sum(!is.na(patient_data$proust_cluster)) > 0) {
  library(mclust)
  
  # Subset to rows with Proust clusters
  patient_data_complete <- patient_data[!is.na(patient_data$proust_cluster),]
  
  # Calculate ARI between original clusterings
  ari_original <- adjustedRandIndex(patient_data_complete$region, 
                                   patient_data_complete$proust_cluster)
  
  # Calculate ARI between merged clusterings
  ari_merged <- adjustedRandIndex(patient_data_complete$merged_clusters, 
                                 patient_data_complete$proust_merged)
  
  cat("\nClustering Method Comparison:\n")
  cat("Adjusted Rand Index (original):", round(ari_original, 3), "\n")
  cat("Adjusted Rand Index (after MMD merging):", round(ari_merged, 3), "\n")
  
  # Visualize cluster overlap
  cluster_overlap <- patient_data_complete %>%
    count(region, proust_cluster) %>%
    group_by(region) %>%
    mutate(prop = n / sum(n)) %>%
    ungroup()
  
  # Plot heatmap of cluster overlap
  ggplot(cluster_overlap, aes(x = region, y = proust_cluster, fill = prop)) +
    geom_tile() +
    scale_fill_viridis_c() +
    theme_minimal() +
    labs(x = "Constrained Hierarchical Clusters", 
         y = "Proust Clusters", 
         fill = "Proportion",
         title = "Cluster Membership Overlap") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
}
```

# Cell Type Composition Analysis

```{r}
# Analyze cell type composition in each cluster
# For standard clustering
cluster_cell_counts <- patient_data %>%
  count(cells, region) %>% 
  rename(count = n)

cell_counts <- patient_data %>%
  count(cells) %>% 
  rename(total_cells = n)

standard_summary <- left_join(cluster_cell_counts, cell_counts, by="cells") %>%
  mutate(proportion = count/total_cells)

# Visualize cell composition in standard clusters
ggplot(standard_summary, aes(x = region, y = proportion, fill = cells)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = cb_palette[1:3]) +
  theme_minimal() +
  labs(x = "Cluster", y = "Proportion", fill = "Cell Type",
       title = "Cell Type Composition in Standard Clusters")

# For Proust clustering (if available)
if (sum(!is.na(patient_data$proust_cluster)) > 0) {
  # Filter out NA values
  proust_data <- patient_data[!is.na(patient_data$proust_cluster),]
  
  proust_cluster_cell_counts <- proust_data %>%
    count(cells, proust_cluster) %>% 
    rename(count = n)
  
  proust_cell_counts <- proust_data %>%
    count(cells) %>% 
    rename(total_cells = n)
  
  proust_summary <- left_join(proust_cluster_cell_counts, proust_cell_counts, by="cells") %>%
    mutate(proportion = count/total_cells)
  
  # Visualize cell composition in Proust clusters
  ggplot(proust_summary, aes(x = proust_cluster, y = proportion, fill = cells)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = cb_palette[1:3]) +
    theme_minimal() +
    labs(x = "Proust Cluster", y = "Proportion", fill = "Cell Type",
         title = "Cell Type Composition in Proust Clusters")
}
```

# Save Results

```{r}
# Save the final plots
ggsave("standard_clusters.png", g_regions, width = 7, height = 5)
ggsave("cell_types.png", g_cells, width = 7, height = 5)
ggsave("merged_clusters.png", g_repart, width = 7, height = 5)

# Save Proust plots if they exist
if (exists("g_proust")) {
  ggsave("proust_clusters.png", g_proust, width = 7, height = 5)
}

if (exists("g_proust_repart")) {
  ggsave("proust_merged_clusters.png", g_proust_repart, width = 7, height = 5)
}

# Save MMD test results
write.csv(standard_test_results, "standard_mmd_tests.csv", row.names = FALSE)

if (exists("proust_test_results")) {
  write.csv(proust_test_results, "proust_mmd_tests.csv", row.names = FALSE)
}
```